\documentclass[]{article}
\usepackage[xcolor]{changebar} 
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{listings}
\newcommand{\deleted}[1]{%
	\cbcolor{red}
	\begin{changebar}
		#1
	\end{changebar}%
}%

\newcommand{\added}[1]{%
	\cbcolor{green}
	\begin{changebar}
		#1
	\end{changebar}%
}%
\usepackage{tikz}
\usetikzlibrary{automata,positioning,arrows}

%opening
\title{Automatic extraction of symbolic network signatures from malware code}
\author{Arjun Gurumurthy, Kausik Subramanian}

\begin{document}

\maketitle

\begin{abstract}
We propose the automatic extraction of symbolic network signatures from malware code to aid and automate network-based malware detection systems.
\end{abstract}

\section{Motivation}
Malware, short for malicious software, 
is one of the primary threats to modern-day networks, especially
with the mass proliferation of the Internet. These are used to 
disrupt computer operations, extract sensitive information, and 
gain access to private systems.  Infected hosts can participate 
in distributed denial-of-service attacks (DDoS), these class of 
attacks are very hard to detect and mitigate, leading to immense
loss of revenues. A new shift is towards creating malware for 
profit called botnets, in which a customer can buy a network of
infected hosts for a price. In such a setting, the malware is
specifically designed to evade detection, thus motivating the need
for better detection tools for malware. 

The most prevalent form of malware detection is endhost-based detection
(commonly known as anti-virus software). Detected malware's code is hashed
and the code signatures are uploaded to a central repository. By maintaining
an up-to-date repository of malware signatures, the endhost software can
detect malware by comparing program signatures with the repository. However, 
endhost-based malware detection can fail due to various reasons~\cite{networksig}:
\begin{itemize}
	\item \textbf{Anti-virus software is not operational or up to date.} 
	Due to this, the endhost machine can be infected by malware without any
	noticeable effects, thus the user will be oblivious of the infection.
	\item \textbf{Packer programs.} To render programs stealthy, malware authors
	employ packer programs~\cite{packer}. Packers change the program content so that
	its signature differs but its functionality has not changed, thus avoiding
	detection by endhost signature-based software.
\end{itemize}

Instead, a intrusion-detection mechanism integrated 
directly into the service provider’s network offers a
much-needed additional layer of protection. While the malware
code is different, the protocol of the malware used to talk to its
control server (which issues orders to the malware) will not vary. 
Thus, by generating accurate \emph{network signature} of the malware (which
can informally defined as the set of messages sent and received by the malware to
initiate an action), we can use the intrusion detection system in the
network (for e.g., Bro~\cite{bro}) to search for these particular signatures,
and can be used to infer the presence of the malware in the network. However,
manually inspecting malware code is an intractable approach (signatures may 
need to be extracted in real-time). Thus, we propose \emph{automatic extraction 
of accurate network signatures from malware code using different static
analysis techniques.} 

\section{Challenges}
We abstract the problem to the following: given a program point,
can we automatically provide an ordered set of symbolic packets 
that must be received (and sent) by the client such that the program 
reaches the program point. The symbolic packets can be used as
network signatures and used to detect malware (a set of packets
satisfying the symbolic packets in order will infer that there 
is a malware with high confidence). By constructing symbolic packet
signatures, we can generalise the intrusion policy and reduce 
true negatives (thus not affecting legitimate traffic).

Malware code can be thought of as a program reacting based on input received
packets received (possibly multi-threaded), which
could require changes to existing reachability analyses tools 
like Codesurfer~\cite{codesurfer}. To extract symbolic network
signatures, we need to define the abstract semantics of the program
suited for network signatures, and define the \emph{combine} and 
\emph{extend} operators used for interprocedural analysis.

\section{Project Proposal}
\textbf{Tasks}:
\begin{itemize}
	\item Identify and characterize features of common malware code - 1-2 weeks.
	Refer to \Cref{sec:malwaremodel}
	\item \added{Extracting a network signature of length $n$ 
		from the program \\
	Refer to \Cref{sec:extraction}
	\item  Generalizing the network signature extraction and further optimizations
	and evaluation.}
	\item \deleted{Modifying/Extending existing tools like Codesurfer for symbolic 
		network signature extraction - 2-3 weeks}
\end{itemize}

\noindent\textbf{Open source tools}: Static analysis tools, malware repositories

\noindent\textbf{Evaluation}: 
The primary metrics of evaluation are time taken for analysis, accuracy and 
generality of the symbolic signatures. Performance analysis can be performed 
with different malware programs with varying lines of code to understand
the time complexity of the analysis. The quality of the symbolic network
signatures is a tougher metric to measure. Ideally, the symbolic network 
signature must only include packets which will make the program reach the
specified point. If the signature is more general, legitimate packets may 
be considered as malicious and blocked. On the other hand, the signature
may miss actual malware traffic because the signature does not encompass
all inputs leading to the point. We would try to address these trade-offs in
our evaluation.

\appendix
\section{Malware program model} \label{sec:malwaremodel}
The malware program listens to incoming packets, and the
network signature is the set of incoming packets which triggers
an attack by the program. We propose a reasonable abstraction
of the code of malware programs. 

We model a malware program as a infinite reactive system 
of the form:
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
}
\end{lstlisting}
Each iteration of the loop, the program receives a
packet, and based on current state and packet received,
it performs certain actions. Using static analysis
techniques, we can identify two set of variables, \emph{state}
and \emph{packet} variables. State variables represent the 
program state in terms of the protocol, which packet variables
are used to represent the current packet being processed by
the program.
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
	if (a == 1 and packet[8] = "b") {
		a = a + 1;
	}
	...
}
\end{lstlisting}	
In the above code snippet, $a$ is a state variable
and $packet$ is a packet variable. This also
illustrates the packet processing 
function of the program, which based on the current
input packet and state variables, either transitions
to a new program state or performs an action (attack or
send packets). 
\begin{lstlisting}	
a = 0;
while(True) {
packet = socket.receive();
	if (a == 0 and packet[0] == "a") {
		a = a + 1;
	}
	if (a == 1 and packet[0] == "b") {
		a = a + 1;
	}
	if (a == 2 and packet[0] == "c") {
		attack()  [the attack point is provided as input to analysis]
	}	
}
\end{lstlisting}	
The above code snippet represents the complete model
of the malware program, where based on input packets and
current state, the program transitions to a new state. 
The symbolic automaton for this program is 
shown in \Cref{fig:progautomaton}. Network signatures
for the malware are all paths from start state ($a=0$)
to state of attack. In this case, the sequence of packets
in order to initiate an attack is $a* \rightarrow b* \rightarrow c*$.
\begin{figure}[h]
	\centering
	\begin{tikzpicture}[shorten >=0.5pt,node distance=3cm,on grid,auto] 
	\node[state, initial] (v0)   {$a = 0$}; 
	\node[state] (v1) [right=of v0] {$a = 1$}; 
	\node[state] (v2) [right=of v1] {$a = 2$};
	\node[state] (v3) [right=of v2] {$attack$}; 	
	\path[->] 
	(v0) edge node {$a*$} (v1)
	(v1) edge node {$b*$} (v2)
	(v2) edge node {$c*$} (v3);
	\end{tikzpicture}
	\caption{Symbolic automaton for malware program.}
	\label{fig:progautomaton}
\end{figure}

\section{Signature Extraction} \label{sec:extraction}
We now describe a procedure to generate a network signature
of length $n$ (length of the signature is the number of packets
received to trigger attack) for a boolean malware program. 
We use the SMT solver Z3~\cite{z3}
to find valid network signatures. In this section, we describe
the constraints generated for the SMT solver from the program
such that the model generated by the solver can be used to 
extract signatures. 

\subsection{Preliminaries}
For a malware program $M$, let $S$ denote
 the vector of boolean state variables in the program,
  and $P$ denote the vector of boolean packet variables
  in the program\footnote{
  	For example, a 60 bit packet is a vector 
  	of 60 boolean variables.}. Let the
  initial values of the state variables be $S_{0}$.

As described in the earlier section, the malware
packet processing code is a set of \texttt{if} 
statements whose predicates are over the set of state
and packet variables. Let the set of \texttt{if} 
statements be denoted by $I$, and let us denote each \texttt{if} 
statement by $(\gamma, \delta) \in I$, where $\gamma(S,P)$
denotes the predicate on $S$ and $P$, and $\delta(S,P)$ denotes
the action on $S$ if $\gamma(S,P)$ is satisfied. Therefore,
the semantics of the \texttt{if} statement is as follows:
\begin{equation}
\gamma(S,P) \Rightarrow S' = \delta(S,P)
\end{equation}
where $S'$ denotes the new state after the \texttt{if} 
block is executed.

\subsection{Transition Model}
We denote the state
vector after receiving $i$ packets as $S_i$.
When packet $i+1$ of the signature is received, 
mlaware $M$, based on $S_i$ and $P_{i+1}$, will execute 
a \texttt{if} $(\gamma_j, \delta_j)$ block\footnote{
	We assume one \texttt{if} block is executed for every packet.}.
Therefore, we can model the malware state transition 
by the following constraints: 
\begin{equation}
	\forall i. 0 \leq i \leq n-2.~\forall (\gamma, \delta) \in I. ~\gamma(S_i, P_{i+1}) \Rightarrow S_{i+1} = \delta(S_i, P_{i+1})
\end{equation}
Let the attack point block be $(\gamma_M, \delta_M)$. 
Therefore, a network signature of length $n$ means
that $P_n$ triggers the attack. The state of the program
is $S_{n-1}$. The following constraint holds:
\begin{equation}
	\gamma_M(S_{n-1},P_{n})
\end{equation}
The SMT solver will return a model for packets $P_1,
\ldots P_n$ such that the program after these receiving
these packets reaches the attack point. Therefore, 
we run this procedure for varying lengths $n$ to find
network signatures of different lengths.

Also, this procedure only returns a particular signature
of length $n$. We can generalize it by analyzing the 
transitions of $M$ to generalize the signature. For
example, if the $i^{th}$ packet only matches certain 
predicates in the transition, we can generalize the packet on
these bits (the rest of the bits are don't care). To 
generate the different singatures of length $n$, 
we can run the SMT solver multiple times to get different 
models satisfying the constraints.

\subsection{Example}
Let us consider the following program receiving two-bit packets:
\begin{lstlisting}	
s1 = 0;
s2 = 0;
while(True) {
	packet = socket.receive();
	if (s1 == 0 and packet[0] == 0) {
	s1 = 1;
	}
	if (s1 == 1 and s2 == 0 and packet[1] == 1) {
	s2 = (not s2) and packet[1];
	}
	if (s1 == 1 and s2 == 1 and packet[0] == 1 and packet[1] == 1) {
	attack()  
	}	
}
\end{lstlisting}	
We denote the state variables by $s_1, s_2$ and packet 
variables by $p_1, p_2$.
For the first \texttt{if} block, 
$\gamma: \neg s_1 \wedge \neg p_1$, and $\delta: s_1'$ ($s_1$ 
is set by the action). Similarily,
for the second block: 
$\gamma: s_1 \wedge \neg s_2 \wedge p_2$, and $\delta: 
s_2' \Leftrightarrow \neg s_2 \wedge p_1$.
For this program, initial state $S_0 = (0, 0)$.
The attack point predicate $\gamma_M$ is as follows:
\begin{equation}
	s_1 \wedge s_2 \wedge p_1 \wedge p_1
\end{equation}
Therefore, if we add the constraints as described in the 
earlier section, for $n=3$, we would obtain a
valid network signature. 
\bibliography{references}

\end{document}
